---
title: "EDA of Titanic Data"
author: "David West"
date: "8/13/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Initial set up of libraries
source(here::here("src","_init_.R"))

# Loading the data set
source(here::here("src","data","make_dataset.R"))

# Feature engineering
source(here::here("src","features","build_features.R"))
```

# Introduction
The purpose of this exploratory analysis is to test out different machine learning algorithms.

# Ground Rules and Assumptions
1. There are NAs for Age in the training data set. A median age will be assumend for all Age NAs. <br />
2. PassengerID, Name, Ticket, and Cabin will be removed from the training data set since there are too many levels for some ML algorithms. May have to come back to this and find a way to incorporate its data still.

```{r}
train2 <- train %>% 
  mutate(
    Age = case_when(
      is.na(Age) ~ median(train$Age, na.rm = T),
      TRUE ~ Age),
    Survived = factor(Survived)) %>% 
  select(-c(Name, PassengerId, Ticket, Cabin))
```


# Models
## randomForest
A randomForest model is created below. 
```{r}
rf_model1 <- randomForest(Survived ~ ., data = train2)

print(rf_model1)
```

## nnet


## svm
## glmnet



